---
title: "Homework 2"
author: Zaynub Ibrahim
output: github_document
---

```{r setup}
library(tidyverse)
library(readxl)
library(dplyr)
```

## Problem 1

Read the Mr.Trashwheel dataset. 

```{r}
trashwheel_df = 
  read_xlsx(
    "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
    sheet = "Mr. Trash Wheel",
    range = cell_cols("A:N")) %>% 
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(
    sports_balls = round(sports_balls),
    sports_balls = as.integer(sports_balls)
  )
```

Read precipitation data for 2018 and 2017. 

```{r}
precip_2018 = 
  read_excel(
    "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
    sheet = "2018 Precipitation",
    skip = 1
  ) %>% 
  janitor::clean_names() %>%
  drop_na(month) %>%
  mutate(year = 2018) %>% 
  relocate(year)

precip_2017 = 
  read_excel(
    "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
    sheet = "2017 Precipitation",
    skip = 1
  ) %>% 
  janitor::clean_names() %>%
  drop_na(month) %>%
  mutate(year = 2017) %>% 
  relocate(year)
```

Now combine annual precipitation. 

```{r}
month_df = 
  tibble(
    month = 1:12,
    month_name = month.name
  )

precip_df = 
  bind_rows(precip_2018, precip_2017)

left_join(precip_df, month_df, by = "month")
```

This dataset contains information from the Mr. Trashwheel trash collector in Baltimore, Maryland. As trash enters the inner harbor, the trashwheel collects that trash, and stores it in a dumpster. The dataset contains information on year, month, and trash collected including some specific kinds of trash. There are a total of `r nrow(trashwheel_df)` rows in our final dataset. Additional data sheets include month precipitation data. 

In 2018, the total precipitation was `r sum(pull(precip_2018, total))`. 

In 2017, the median number of sports balls in a dumpster was `r median(pull(filter(trashwheel_df, year == 2017), sports_balls))`.

## Problem 2

Read the NYC Transit dataset. Clean data, retain specific variables, and convert variable entry from character to logic. 

```{r}
transit_data = read_csv(file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
      janitor::clean_names() %>%
   select(line:entry, vending, ada) %>%
   mutate(entry = recode(entry, "YES" = "TRUE", "NO" = "FALSE")
   )
```

This dataset contains `r nrow(transit_data)` observations with `r nrow(transit_data)` rows and `r ncol(transit_data)` columns. Variables contained in this dataset are `r names(transit_data)`. The cleaning steps taken so far include changing variable names to snake case, keeping only specific variables using the select function, and changing the entry variable from a character variable to a logical variable using the mutate function.  

However, despite this cleaning, the data is not tidy yet since each route is its own variable and some are listed as character variable while others are double vector variables. We need to convert them all to character and then merge into one variable to make it easier to work with, using the pivot_longer function. 

```{r}
transit_tidy = transit_data %>%
  mutate( 
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
    ) 

transit_tidy = 
  pivot_longer(
    transit_tidy, 
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    names_prefix = "route"
    )
```

This tidy dataset contains `r nrow(transit_tidy)` observations with `r nrow(transit_tidy)` rows and `r ncol(transit_tidy)` columns. Variables contained in this dataset are `r names(transit_tidy)`. 

In this dataset, there are `r distinct(transit_tidy, line, station_name) %>% count` distinct stations. Among theses, `r filter(transit_tidy, ada == "TRUE") %>% distinct(line, station_name) %>% count` are ADA compliant. The proportion of station entrances and exits without vending that allow entrance are `r filter(transit_tidy, entry == "TRUE", vending == "NO") %>% count()/filter(transit_tidy, vending == "NO") %>% count()`. 

There are `r filter (transit_tidy, route_name == "A") %>% distinct(line, station_name) %>% count()` distinct stations that seve the A train and there are `r filter (transit_tidy, route_name == "A") %>% filter(ada == "TRUE") %>% distinct(line, station_name) %>% count()` distinct stations that both serve the A train and are ADA compliant. 


## Problem 3

*Clean the pols-month.csv dataset.* 
Separate the mon variable into three new variables labeled year, month, and day. Create a new president variable and remove prez_gop, prez_dem, and day variables.   

```{r}
pols_month = read_csv(file = "./data/pols-month.csv") %>%
      janitor::clean_names() %>% 
    separate(
      mon,
      c("year", "month", "day")) %>% 
    mutate(
      month = month.name[as.factor(month)],
      year = as.factor(year),
      president = case_when(
        prez_gop == 1 ~ "gop",
        prez_dem == 1 ~ "dem")) %>%
      select(
        -c(prez_dem, prez_gop, day)) %>%
    relocate(year, month)
```

*Clean the snp.csv dataset*
Separate the date into month and year, change variable types, remove day variable. 

```{r}
snp = read_csv(file = "./data/snp.csv") %>%
      janitor::clean_names() %>% 
    separate(
      date,
      c("year", "month", "day")) %>% 
    mutate(
      month = month.name[as.factor(month)],
      year = as.factor(year)) %>% 
    select (-c(day)) %>%   
    relocate(year, month)
```

*Tidy the unemployment data* 







